### 2024-07-11 causal-inference.qmd
20240711 - multiple figures

One of the plausible alternatives to the inference that `X` causes `Y` is that there are third variable [confounds](#sec-causalDiagramConfounding) that influence both `X` and `Y`, thus explaining why `X` and `Y` are associated, as depicted in @fig-correlationAndCausation3 and @fig-ZCausesXandY.
Thus, another approach that can help increase [internal validity](#sec-internalValidity) is to include plausible [confounds](#sec-causalDiagramConfounding) as control variables.
For instance, if a third variable such as education level might be a [confound](#sec-causalDiagramConfounding) that influences both sports drink consumption and player performance, you could include education level as a [covariate](#sec-covariates) in the model.
Inclusion of a [covariate](#sec-covariates) attempts to control for the variable by examining the association between the [predictor variable](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy) while holding the [covariate](#sec-covariates) variables constant.
For instance, such a model would examine whether, when accounting for education level, there is an association between sports drink consumption and player performance.

@@

A key tool when describing a research question or hypothesis is to create a conceptual depiction of the hypothesized causal processes.
A causal diagram depicts the hypothesized causal processes that link two or more variables.
A common form of causal diagrams is the directed acyclic graph (DAG).
DAGs provide a helpful tool to communicate about causal questions and help identify how to avoid bias (i.e., over-estimation) in associations between variables due to [confounding](#sec-causalDiagramConfounding) (i.e., common causes) [@Digitale2022].
For instance, from a DAG, it is possible to determine what variables it is important to control for in order to get unbiased estimates of the association between two variables of interest.
To create DAGs, you can use the `R` package `dagitty` [@Textor2017] or the associated browser-based extension, DAGitty: <https://dagitty.net> (archived at <https://perma.cc/U9BY-VZE2>).
Examples of various causal diagrams that could explain why `X` is associated with `Y` are in @fig-XCausesY, @fig-YCausesX, and @fig-ZCausesXandY.

@@

*Collision* occurs when two variables influence a third variable, the collider [@DOnofrio2020].
That is, a collider is a variable that is caused by two other variables (i.e., a common effect).
An example collision is depicted in @fig-colliderBias1 and @fig-colliderBias2:

>>>>

One of the plausible alternatives to the inference that `X` causes `Y` is that there are third variable [confounds](#sec-causalDiagramConfounding) that influence both `X` and `Y`, thus explaining why `X` and `Y` are associated, as depicted in Figures [-@fig-correlationAndCausation3] and [-@fig-ZCausesXandY].
Thus, another approach that can help increase [internal validity](#sec-internalValidity) is to include plausible [confounds](#sec-causalDiagramConfounding) as control variables.
For instance, if a third variable such as education level might be a [confound](#sec-causalDiagramConfounding) that influences both sports drink consumption and player performance, you could include education level as a [covariate](#sec-covariates) in the model.
Inclusion of a [covariate](#sec-covariates) attempts to control for the variable by examining the association between the [predictor variable](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy) while holding the [covariate](#sec-covariates) variables constant.
For instance, such a model would examine whether, when accounting for education level, there is an association between sports drink consumption and player performance.

@@

A key tool when describing a research question or hypothesis is to create a conceptual depiction of the hypothesized causal processes.
A causal diagram depicts the hypothesized causal processes that link two or more variables.
A common form of causal diagrams is the directed acyclic graph (DAG).
DAGs provide a helpful tool to communicate about causal questions and help identify how to avoid bias (i.e., over-estimation) in associations between variables due to [confounding](#sec-causalDiagramConfounding) (i.e., common causes) [@Digitale2022].
For instance, from a DAG, it is possible to determine what variables it is important to control for in order to get unbiased estimates of the association between two variables of interest.
To create DAGs, you can use the `R` package `dagitty` [@Textor2017] or the associated browser-based extension, DAGitty: <https://dagitty.net> (archived at <https://perma.cc/U9BY-VZE2>).
Examples of various causal diagrams that could explain why `X` is associated with `Y` are in Figures [-@fig-XCausesY], [-@fig-YCausesX] and [-@fig-ZCausesXandY].

@@

*Collision* occurs when two variables influence a third variable, the collider [@DOnofrio2020].
That is, a collider is a variable that is caused by two other variables (i.e., a common effect).
An example collision is depicted in Figures [-@fig-colliderBias1] and [-@fig-colliderBias2]:

### 2024-07-11 research-methods.qmd
20240711 - multiple figures

For example, if the effect of consuming sports drinks on player performance differs for Quarterbacks and Wide Receivers, the interaction could be depicted in @fig-moderationExamplePathDiagram and @fig-moderationExampleInteractionGraph:

>>>>

For example, if the effect of consuming sports drinks on player performance differs for Quarterbacks and Wide Receivers, the interaction could be depicted in Figures [-@fig-moderationExamplePathDiagram] and [-@fig-moderationExampleInteractionGraph]:
